---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
```{r echo=TRUE}
download.file(url="https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", destfile="data.zip")
unzip("data.zip")
df = read.csv("activity.csv")
df$date = as.Date(df$date, "%Y-%m-%d")
str(df)
```

## Histogram of the total number of steps taken each day
```{r echo=TRUE,message=FALSE}
library(dplyr)
d_steps = df %>% group_by(date) %>% summarize(daily_steps=sum(steps)) #this gives the total nb of steps per day
hist(d_steps$daily_steps, labels=TRUE, lwd=10, col=10,xlab="Steps", main="Total nb of daily steps per day")
```

## What is mean and the median of the total number of steps taken per day?
```{r echo=TRUE,message=FALSE}
avg = as.integer(mean(d_steps$daily_steps, na.rm=TRUE))
med = median(d_steps$daily_steps, na.rm=TRUE)
```
The mean of the total nb of steps per day = `r avg` and the median = `r med`

## What is the average daily activity pattern?
```{r echo=TRUE,message=FALSE}
i_steps = df %>% group_by(interval) %>% summarize(avg=mean(steps, na.rm=TRUE))
with(i_steps, plot(interval, avg, type="l"))
title(main = "Avg nb of steps by interval")
```

## Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r echo=TRUE,message=FALSE}
max = as.integer(i_steps[which.max(i_steps$avg),1])
``` 
The 5-min interval that contains (on avg) the maximum nb of steps is the one identified by `r max`

## Imputing missing values
```{r echo=TRUE,message=FALSE}
na_count = sum(is.na(df$steps))
#Let's assign the mean for that 5-minute interval to the missing values
intervals = unique(df[,"interval"])
for (i in intervals){
  i_avg = as.integer(subset(i_steps, interval==i)["avg"]) #get the avg nb of steps for this interval
  df$steps[is.na(df$steps) & df$interval==i] = i_avg
}
head(df)
```

## Histogram of the total number of steps taken each day with the new datasets (without NAs)
```{r echo=TRUE,message=FALSE}
d_steps = df %>% group_by(date) %>% summarize(daily_steps=sum(steps)) #this gives the total nb of steps per day
hist(d_steps$daily_steps, labels=TRUE, lwd=10, col=10,xlab="Steps", main="Total nb of daily steps per day (new dataset w/o NAs)")
```

If we compare the 2 histograms (the one before and the one after inputting missing values), the "shape" remains the same. The only difference is: the nb of days where the step count is btw 10.000 and 15.000 has increased by 8 (before 28, now 36) 

## Mean and median total number of steps taken per day with the new datasets (without NAs)
```{r echo=TRUE,message=FALSE}
avg = as.integer(mean(d_steps$daily_steps, na.rm=TRUE))
med = median(d_steps$daily_steps, na.rm=TRUE)
```
The mean of the total nb of steps per day = `r avg` and the median = `r med`. 
There is a minor difference with the mean and median we had before changing the NA values


## Are there differences in activity patterns between weekdays and weekends?
```{r echo=TRUE,message=FALSE}
# Let's create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
df$week = ifelse(weekdays(df$date) %in% c("Saturday", "Sunday","samedi","dimanche"), "weekend", "weekday")
df$week = as.factor(df$week)
str(df)

```

```{r echo=TRUE,message=FALSE}
# Let's create a plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
library(lattice)
i_steps = df %>% group_by(interval,week) %>% summarize(avg=mean(steps, na.rm=TRUE))
xyplot(avg~interval | week, data=i_steps, layout=c(1,2), type="l")
```